Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-26 write-ups/FSE/paper.tex
A little negative spacing.
Remove mention of supplementary material for semantics and other design choices; this isn't so important.
More space.

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t) {
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
    }
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
%\vspace{-6mm}
\end{figure}

@@

\begin{lstlisting}
kernel graph_app(global graph * g,
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;
  int tid = get_global_id();
  int stride = get_global_size();
(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
%\vspace{-2mm}
\end{figure}

@@

We present our cooperative kernels programming model as an extension
to OpenCL.  However, the cooperative kernels concept is more general,
and could be applied to extend other GPU programming models, such as
CUDA and HSA.  We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}).
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%
In Supplementary material we present a formal semantics for
cooperative kernels atop an abstract GPU programming model, and
discuss several cases where we might have taken different and also
reasonable semantic decisions.

@@

%\vspace{-1mm}
\myparagraph{Transmitted variables}
%
A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  Annotating
a variable $v$ with $\transmit$ means that when a workgroup spawns new
workgroups by calling $\offerfork$, the workgroup should transmit its
current value for $v$ to the threads of the new workgroups, to serve
as an initial value for $v$.  We detail the semantics for this when we
describe $\offerfork$ below.

@@

When executed by a cooperative
kernel, $\getnumgroups$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time (see \mysec\ref{sec:opencl}).

@@

Suppose a workgroup with id $m$ executes $\offerkill$.  If the
workgroup has the highest largest id of all active workgroups then it
can be killed by the scheduler, except that workgroup 0 can never be
killed (to ensure that the kernel computation does not terminate
early).  More formally, if $m < M-1$ or $M=1$ then $\offerkill$ is a
no-op.  If instead $M > 1$ and $m = M-1$, the scheduler can choose to
ignore the offer, so that $\offerkill$ executes as a no-op, or accept
the offer, so that execution of the workgroup ceases and the number of
active workgroups $M$ is atomically decremented by one.
\myfiglong{\ref{fig:overview}} illustrates this, showing that
workgroup $3$ is killed before workgroup $2$.
%In the second case, the scheduler can assign some other task to
%the compute unit on which the workgroup was executing.

@@

\myparagraph{Launching kernels and managing workgroups}
%
To launch a kernel, the host sends a data packet to the GPU scheduler
consisting of: a kernel to execute, the kernel inputs, and a flag
indicating whether the kernel is cooperative. In our implementation,
this host-device communication channel is built using fine-grained SVM
atomics.

@@

Among the GPUs available to us, four met our requirements: Intel HD
520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
drivers for OpenCL suffer from a known defect whereby long-running
kernels lead to unkillable defunct
processes~\cite{amd-defunct-process}. Because of this behaviour, we
could not experiment on AMD platforms.

%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.

We thus ran our experimental campaign on the Intel GPUs. The results
were similar across the GPUs, so for conciseness, we report only on
the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
Iris has a reported 47 compute units. Results for the two other Intel
GPUs are available in supplementary material.

@@

The low overhead of cooperative kernels for low and medium graphics
workloads is due to sub-linear scalability of the irregular algorithms that they implement: some resources can
be taken away during computation without proportionally reducing
execution time.  Our poor performance for the heavy workload is
because the graphics task requires all but one workgroup, which
cripples the cooperative kernel enough that $\offerfork$ calls are not
issued as frequently.

These results suggest that a hybrid preemption scheme may work the
best for multi-tasking and graphics. That is, the cooperative approach
works well for light and medium tasks, but heavy graphics tasks
benefit from the coarser grained, kernel-level preemption strategy.

>>>>

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t) {
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
    }
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
%\vspace{-6mm}
\end{figure}

@@

\begin{lstlisting}
kernel graph_app(global graph * g,
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;
  int tid = get_global_id();
  int stride = get_global_size();
(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
%\vspace{-2mm}
\end{figure}

@@

We present our cooperative kernels programming model as an extension
to OpenCL.  However, the cooperative kernels concept is more general,
and could be applied to extend other GPU programming models, such as
CUDA and HSA.  We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}).
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%

@@

%\vspace{-1mm}
\myparagraph{Transmitted variables}
%
A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  Annotating
a variable $v$ with $\transmit$ means that when a workgroup 
uses $\offerfork$ to spawn new
workgroups, the workgroup should transmit its
current value for $v$ to the threads of the new workgroups.
We detail the semantics for this when we
describe $\offerfork$ below.

@@

When executed by a cooperative
kernel, $\getnumgroups$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups$ for
regular kernels, which returns the fixed number of workgroups that execute the kernel (see \mysec\ref{sec:opencl}).

@@

Suppose a workgroup with id $m$ executes $\offerkill$.  If the
workgroup has the largest id among active workgroups then it
can be killed by the scheduler, except that workgroup 0 can never be
killed (to avoid early termination of the kernel).  More formally, if $m < M-1$ or $M=1$ then $\offerkill$ is a
no-op.  If instead $M > 1$ and $m = M-1$, the scheduler can choose to
ignore the offer, so that $\offerkill$ executes as a no-op, or accept
the offer, so that execution of the workgroup ceases and the number of
active workgroups $M$ is atomically decremented by one.
\myfiglong{\ref{fig:overview}} illustrates this, showing that
workgroup $3$ is killed before workgroup $2$.
%In the second case, the scheduler can assign some other task to
%the compute unit on which the workgroup was executing.

@@

\myparagraph{Launching kernels and managing workgroups}
%
To launch a kernel, the host sends a data packet to the GPU scheduler
consisting of a kernel to execute, kernel inputs, and a flag
indicating whether the kernel is cooperative. In our prototype,
this host-device communication channel is built using fine-grained SVM
atomics.

@@

Among the GPUs available to us, four met our requirements: Intel HD
520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
drivers for OpenCL suffer from a known defect whereby long-running
kernels lead to unkillable defunct
processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
We thus ran our experimental campaign on the Intel GPUs. The results
were similar across the GPUs, so for conciseness, we report only on
the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
Iris has a reported 47 compute units. Results for the two other Intel
GPUs are available in supplementary material.

@@

The low overhead of cooperative kernels for low and medium graphics
workloads is due to sub-linear scalability of the irregular algorithms that they implement: some resources can
be taken away during computation without proportionally reducing
execution time.  Our poor performance for the heavy workload is
because the graphics task requires all but one workgroup, which
cripples the cooperative kernel enough that $\offerfork$ calls are not
issued as frequently.
%
These results suggest that a hybrid preemption scheme may work the
best for multi-tasking and graphics. That is, the cooperative approach
works well for light and medium tasks, but heavy graphics tasks
benefit from the coarser grained, kernel-level preemption strategy.
